---
title: "Modelling Radionuclide Concentrations in Fucus on the Swedish West Coast"
subtitle: "A Bayesian Distributed Lag Approach"
title-block-banner: true
affiliation-title: "Division of Particle and Nuclear Physics\nDepartment of Physics\nLund Unversity"
author: "Pim Nelissen"
date: 2026-01-14
format: html
editor: visual
embed-resources: true
toc: true
---

::: callout-note
To reproduce/run this notebook yourself you can clone the following repository:

```         
https://github.com/pim-n/sardal-radioactivity-analysis
```
:::

## 1. Introduction

This is the Quarto markdown document that will go through the entire data analysis process done for this project. Most code is called in from *helper* files, to ensure the document remains readable. As functions are called in from these helper files, their workings will be explained. In case you are interested in seeing the source code, refer to the helper files in the `R/` folder at the [GitHub repository](https://github.com/pim-n/sardal-radioactivity-analysis/tree/main/R).

We will now load these helper files:

```{r}
source("R/data_helpers.R")
source("R/model_helpers.R")
source("R/plotting_helpers.R")
```

**Section 2** of this document will deal with the data wrangling process. **Section 3** will cover data exploration and visualization. **Section 4** will give an example of how a model is specified, and **Section 5** covers results and analysis of all isotopes.

For mathematical and theoretical details, refer to the project report.

### Acknowledgment

The work here is centered around the `mvgam` \[1\] package, an R package for Bayesian fitting of Dynamic Generalized Additive Models. Some code is adapted and modified from various sources. In case code has been adapted from somewhere, this is specified in the corresponding section of the document.

\[1\] Clark N, Wells K (2023). “Dynamic Generalized Additive Models (DGAMs) for forecasting discrete ecological time series.” *Methods in Ecology and Evolution*, **14**, 771-784. [doi:10.18637/jss.v100.i05](https://doi.org/10.18637/jss.v100.i05).

### Running option

If below variable is TRUE, run the entire document including creating the models. Otherwise, load the models from `results/models/`.

```{r}
#| echo: true
#| output: false

FULL_RUN = FALSE
MIN_LAG = 1
MAX_LAG = 8

if (!FULL_RUN) {
  model_files <- paste0("results/models/", list.files("results/models"))
  lapply(model_files, load, environment())
}
```

### Importing libraries

The following packages were used in this notebook:

```{r}
#| echo: true
#| output: false

# data wrangling
library(tidyverse)
library(glue)
library(lubridate)
library(readxl)

# interpolation of time series
library(zoo)

# for fitting the distributed lag GAMs
library(mgcv)
library(mvgam)

# plotting and formatting
library(gratia)
library(viridis)
```

## 2. Data Wrangling

> **Summary**: An explanation of the pre-processing procedure, followed by import of CSVs into R dataframes compatible with the `tidy` principles.

### Pre-processing

The data was provided as a set of Excel `.xlsx` files with sheets for the various isotopes. The data was manually put in CSV format, which is an open standard which is suitable for later import into R. The choice was made to separate the data by type (discharge or measurement at Särdal) and isotope. For measurements, a further distinction was made between species of Fucus *spp.*. The structure of the pre-processed data is shown below:

``` bash
.
├── discharges
│   ├── discharges_C14.csv
│   ├── discharges_Co60.csv
│   ├── ...
└── sardal
    ├── C14
    │   └── sardal_activity_C14_F_Various_Mattsson_2025.csv
    ├── Co60
    │   ├── sardal_activity_Co60_F_Serratus_Mattsson_2025.csv
    │   └── sardal_activity_Co60_F_Vesiculosus_Mattsson_2025.csv
    ├── Cs134
    │   ├── ...
    ├── ...
    └── ...
```

### Data loading

Now it's time to load the CSVs into a usable format for data-processing. The process is as follows

1.  List all isotope CSV files in the Särdal path
2.  For each CSV, map to a dataframe, setting column per species of Fucus
3.  Find the earliest and latest observation date across all dataframes, and create a `obs_date` sequence from the earliest to latest sample date
4.  Join together all the isotope dataframes to this unified `obs_date` column, giving one sardal dataframe `df_sardal`

```{r}
#| echo: true
#| output: false

sardal_path <- "data/sardal" 
sardal_files <- list.files(sardal_path, full.names = TRUE, pattern = "*.csv", recursive = TRUE) %>% set_names()
sardal_df_list <- purrr::map(sardal_files, read_csv,
                  col_types = cols(date = col_date(format = "%Y-%m-%d")))

sardal_df_list <- map2(sardal_df_list, names(sardal_df_list),
     function(.x, .y){dplyr::select(.x, 1:2) %>%
         set_names(nm=c("obs_date", get_species_name(.y)))}) %>%
          map(~drop_na(.x))

df_sardal <- data.frame(obs_date = get_full_time_range_by_day(sardal_df_list))
df_sardal <- map(sardal_df_list, ~ full_join(df_sardal, .x, by = "obs_date"))
df_sardal <- reduce(df_sardal, full_join, by = "obs_date")

df_long_sardal <- df_sardal %>%
  pivot_longer(
    cols = -obs_date,
    names_to = "variable",
    values_to = "value"
  )
```

The same for steps are applied for the discharges, except we have no species and have a `year` time column instead of `obs_date`.

```{r}
#| echo: true
#| output: false

discharges_path <- "data/discharges" 
discharges_files <- list.files(discharges_path, full.names = TRUE, pattern = "*.csv", recursive = TRUE) %>% set_names()
discharges_df_list <- purrr::map(discharges_files, read_csv)

discharges_df_list <- map2(discharges_df_list, names(discharges_df_list),
                  function(.x, .y) {
                    isotope_name <- get_isotope_name(.y)
                    column_names <- colnames(.x)[-1] # exclude year col.
                    new_column_names <- c(colnames(.x)[1], paste(isotope_name, column_names, sep = "_")) # concatenate isotope name to cols.
                    
                    .x %>%
                      set_names(nm = new_column_names) # update col. names
                  }) %>%
  map(~drop_na(.x))

df_discharges <- data.frame(year = get_full_time_range_by_year(discharges_df_list))
df_discharges <- map(discharges_df_list, ~ full_join(df_discharges, .x, by = "year"))
df_discharges <- reduce(df_discharges, full_join, by = "year")

df_long_discharges <- df_discharges %>%
  pivot_longer(
    cols = -year,
    names_to = "variable",
    values_to = "value"
  )
```

## 3. Data exploration & visualization

> **Summary**: An explanation of the data through visualization. Data quality is discussed by looking at amount of data available and missing values. Finally, an autocorrelation test is performed which shows high autocorrelation.

### Plotting data

```{r}
data_Cs137 <- combine_data("Cs137", df_long_sardal, df_long_discharges)
p <- plot_isotope(data_Cs137)
p
```

or Co60:

```{r}
data_Co60 <- combine_data("Co60", df_long_sardal, df_long_discharges)
p <- plot_isotope(data_Co60)
p
```

### Number of observations

For statistical analysis it is very important to know how much data one has, and how it is structured, and whether there are any missing values or irregularities. A first step could be to see how many unique daily measurements and unique yearly measurements we have:

```{r}
sum_daily <- df_sardal %>%
  group_by(obs_date) %>%
  summarize(across(where(is.numeric), ~ any(!is.na(.x)))) %>%
  summarize(across(-obs_date, sum)) %>%
  pivot_longer(everything())

sum_yearly <- df_sardal %>%
  mutate(year = year(obs_date)) %>%
  group_by(year) %>%
  summarize(across(where(is.numeric), ~ any(!is.na(.x)))) %>%
  summarize(across(-year, sum)) %>%
  pivot_longer(everything())

sums_table <- merge(sum_daily, sum_yearly, by = 'name') %>%
  rename(days = value.x, years = value.y)
      
sums_table
```

Here is a visualization of the above information, showing the gaps in measured activity concentration (on a yearly scale):

```{r}
df_long_sardal %>%
  filter(!(variable %in% c("I129_F_Serratus_Winter", "I129_F_Serratus_Summer"))) %>%
  plot_na_values()
```

### Autocorrelation

Autocorrelation is a measure of correlation of a time series with a lagged (delayed) version of itself. This means it can quantify if, and how much, past values affect the present value.

::: callout-note
Output is suppressed for the autocorrelation plots because there are many. To see them, consult the written report or clone the repo and run this notebook.
:::

```{r}
#| echo: true
#| output: false

lapply(names(df_discharges)[-1], function(col) {
  col_data <- na.omit(df_discharges[[col]])
  acf(col_data, plot = TRUE, main = paste("ACF Plot for", format_site_label(col)))
})
```

```{r}
#| echo: true
#| output: false

lapply(names(df_sardal)[-1], function(col) {
  col_data <- na.omit(df_sardal[[col]])
  acf(col_data, plot = TRUE, main = paste("ACF Plot for", format_species_label(col)))
})
```

We can see that there is some autocorrelation for almost all data. However, explicitly adding an AR component is not always useful. The best case to apply it is Carbon-14, since there is a known external effect (the bomb pulse) This means we have to somehow capture this "memory" of past values in the model. One way is to add an autoregressive component. This will be done in the model specification.

### Adressing missing values in predictors

Now, we will combine the discharge and activity concentration measurements for a given isotope, so that we can use that dataframe in the model.

Due to limitations in the `mvgam` package, predictors **cannot** have NaN values. Therefore, where missing years are present in the discharges we must address this. Let's see if we have this issue:

```{r}
plot_na_values(df_long_discharges)
```

We can see that for the caesium isotopes, there is some years without any discharge data available. We will therefore interpolate the data in these gaps in order to make fitting a model possible for caesium as well. The `interpolate_data` function will take care of this.

Now, let's plot an example and see if the interpolation looks reasonable:

```{r}
vars <- df_long_discharges %>%
  filter(str_detect(variable, "Cs137")) %>%
  distinct(variable)

# Uninterpolated data
uninterp <- df_long_discharges %>%
  filter(str_detect(variable, "Cs137"),
         year > 2010) %>%
  mutate(type = "Uninterpolated")

# Interpolated ONCE using fuzzy isotope string
interp <- interpolate_data("Cs137", df_long_discharges) %>%
  filter(year > 2010) %>%
  mutate(type = "Interpolated")


# Combine
plot_df <- bind_rows(interp, uninterp)

# Plot
ggplot(plot_df, aes(x = year, y = value, color = type)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    x = "Year",
    y = "Discharge [TBq/y]",
    color = "Data type"
  ) +
  theme_minimal()
```

Looks fine!

## 4. Example modelling process: Carbon-14

> **Summary**: The chapter goes through creating an isotope dataset with the hand of carbon-14. Discharge and Särdal measurements are combined into one isotope-specific dataset. Tensor smooths are introduced as the foundational component for the distributed lag model.

### Combining measurements and discharge

Now it is time to combine the data. The `combine_data` function will, for a given isotope string (e.g. "C14"), combine the Särdal measurement data as well as the discharge data for said isotope, returning one comprehensive dataframe.

```{r}
data_C14 <- combine_data("C14", df_long_sardal, df_long_discharges, scale_data = FALSE)
plot_isotope(data_C14)
```

One thing that may be relevant is the correlation between Sellafield and La Hague here. This is to avoid co-linearity issues in the model. Basically, if the time series are highly correlated, adding them as separate predictors in a regression model can lead to inaccurate results. To do this, we must first make the series quasi-stationary by taking the first difference, before checking the correlation.

```{r}
cor(diff(data_C14$sellafield), diff(data_C14$la_hague))
```

This is not a significant value to worry about, so we do not take any special steps to avoid co-linearity in the predictors.

### Lag matrix

> See Wood (2017) - *Generalized Additive Models: An Introduction with R*, section 5.6 for a detailed explanation on tensor smooths.

The lag matrix is a matrix of the form

```         
0 1 2 3 4 ...
0 1 2 3 4 ...
0 1 2 3 4 ...
0 1 2 3 4 ...
...
```

with the dimensions of $N \times L$, where $N$ is the number of data points and $L$ is the maximum lag. This construction is needed in order to create a tensor smooth. A tensor product smooth is a smooth surface that is built up from several smooth basis functions. In this case, we create a smooth surface varying with discharge amount from source $x$ and the lag $l$, where the height of the surface represents the effect on $y_t^{sp}$ at a given time $t$

$$
f(x_{t-l}, l) = \sum_{k=0}^K \sum_{m=0}^M \beta_{km} b_k(x_{t-l}) c_m(l)
$$

here,

-   $b_k(x_{t-l})$ are the basis functions in discharge amount $x_{t-l}$

-   $c_m(l)$ are the basis functions in lag $l$

and $K$ and $M$ specify the allowed complexity (number of terms) that are allowed to represent those basis functions. For more details, see Wood (2017) and the written report.

```{r}
data_C14 <- data_C14 %>%
  add_lagged_matrices(n_max = MAX_LAG, n_min = MIN_LAG)
```

### Hierarchical structure

A hierarchical model is useful when you have a dataset which consists of distinct groups that may have varying effects. Let's explain by example of our Fucus *spp.*.

One could create two seperate models for Fucus *Serratus* and Fucus *Vesiculosus* respectively, modelling the effect of the discharges on each of them individually. But here we ignore the fact that there are common effects among them, namely the transport mechanisms that bring radionuclides to the Swedish coast. Instead of modelling them separately

$$
\text{Activity in F. Serratus} \sim \text{Discharge}\\
\text{Activity in F. Vesiculosus} \sim \text{Discharge}
$$

One creates a group intercept

$$
\text{Activity} \sim (1 \vert \text{species}) + \text{Discharge}
$$

We essentially add a dummy variable to allow variable intercept that depends on the discharge.

measured activity concentrations, or doing something arbitrary like taking the average, a *hierarchical* modelling structure combines the information by taking both datasets in the regression, but allowing some statistical variation between them.

To achieve this, we apply weight or 'mask' matrices that are the size of the lag matrices, but consist of zeroes and ones. Simply, `weights_s` matrix is 1 when the species is F. Serratus, 0 otherwise, and the same is true for the F. Vesiculosus weight matrix `weights_v`.

Note: We are not actually going to use these for the Carbon-14 model, but the procedure is just explained by the hand of C-14.

```{r}
data_C14 <- data_C14 %>%
  apply_weights()
```

### Mathematical formulation

With the functions defined and explained, we now specify the model:

$$
g(y_t^*) = \underbrace{\alpha}_{\text{intercept}} + \underbrace{\sum_{l=0}^L f(\text{SF}_{t-l},l)}_{\text{Sellafield contribution}} + \underbrace{\sum_{l=0}^L f(\text{LH}_{t-l},l)}_{\text{La Hague contribution}} + \underbrace{u_t}_{\text{AR(1)}} + \underbrace{\varepsilon_t}_{\text{error term}}
$$

where $u_t = \rho u_{t-1}$ and

-   $y_t^{*}$ is bomb-pulse corrected time series $y_t - 0.68 \cdot y_t^{bomb}$
-   $l \in \{0, L\}$ is the lag in years.
-   $g(\cdot)$ is the link function, that links the linear combination of predictors to the expected value of $y_t^{sp}$.
-   $f(x_{t-l},l)$ are the tensor product smooths as defined before.
-   $\varepsilon_t \sim \mathcal{N}(0, \sigma^2)$ is the error term.

### Fitting the model

We can now fit the model. Note log-link relation between the predictors and response (for more details on why, see report).

The fitting can now be done by first formulating the model in mgcv/mvgam convention, then calling the functions. Note that the family of response is now assumed to be $y_t \sim \text{LogNormal}(\mu, \sigma^2)$, which in terms of distribution of $y_t$ is equivalent to $log(y_t) \sim \text{Normal}(\mu, \sigma^2)$, with the small benefit that we do not have to transform the data back to the proper scale after fitting.

#### C14 Model - AR(1)

```{r}
#| echo: true
#| output: false

if(FULL_RUN) {
  formula <- sardal ~
    te(sellafield, lag, k=4) +
    te(la_hague, lag, k=4)
  
  priors <- get_mvgam_priors(
    formula = formula,
    family = lognormal(),
    data = data_C14,
    trend_model = AR(1)
  )
  
  priors$prior[3] <- 'ar1 ~ beta(2, 2);'
  
  model_C14_AR_mvgam <- mvgam(formula = formula,
                           family = lognormal(),
                           data = data_C14,
                           trend_model = AR(1),
                           burnin = 2000,
                           samples = 4000,
                           parallel = TRUE,
                           threads = 6,
                           noncentered = TRUE,
                           priors = priors
                           )
  
  save(model_C14_AR_mvgam, file = "results/models/C14_AR.mvgam.Rdata")
}
```

And that's it! The `C14.mvgam.Rdata*` is now a complete fitted model object containing estimated parameters, original data, predicted (fitted) data, STAN code, posterior information and more. We save the model.

## 6. All results

### Carbon-14

Before we fit the other models we will now look at the summary of objects created above

```{r}
summary(model_C14_AR_mvgam)
```

```{r}
mcmc_plot(model_C14_AR_mvgam, variable='ar1', regex=TRUE, type="areas")
```

We see that the AR(1) component is very strong, indicating a strong temporal trend. This is the residual of the carbon-14 bomb pulse that is being captured by the AR(1) trend component. The tensor product smooth posteriors are a bit difficult to interpret directly. Instead, we can look at the final estimated interaction effect surface from the fitted tensor product smooths:

```{r}
plot_mvgam_smooth(model_C14_AR_mvgam, series = 1, smooth = 1)
plot_mvgam_smooth(model_C14_AR_mvgam, series = 1, smooth = 2)
```

Again, these surfaces represent $f(x_{t-l}, l)$, essentially the 'joint effect' of lag and discharge on the measured activity concentration in Fucus at Särdal. Sellafield is showing a pattern in line with expectations: a higher discharge from Sellafield results in a higher (more positive) effect on Särdal activity concentrations. Regarding lag, we see that the highest values are centered around 3-5 years, with a very broad delayed effect up to 8 years as well, with the effect becoming closer to 0 near 9-10 years. This seems in line with estimates of Sellafield transport time being approximately 3-4 years, corresponding to the initial 'hit', and the prolonged effect across many lags could be due to the delayed uptake as other environmental processes (which? ask Kristina). La Hague does not really match expectation.

We can also inspect what the predictions are from this fitted model (note the log scale!):

```{r}
plot(hindcast(model_C14_AR_mvgam))
```

Some model diagnostics that will inform us about the fit are the residual plots and the prior-posterior check:

```{r}
plot_mvgam_resids(model_C14_AR_mvgam)
ppc(model_C14_AR_mvgam)
```

The residual plots, such as the Q-Q plot, indicate a good model fit, suggesting that the lognormal family is appropriately describing the response. There is some slight residual autocorrelation which is not captured by the AR(1) process, overall the AR(1) has proved effective at capturing both the effect of the bomb pulse and the 'memory' effect of radiocarbon being stored in the Fucus over time.

The posterior predictive check (PPC) shows how well the estimated posterior $\hat{y}$ represents the distribution of the observed data $y$. We can see a rather good fit. However, because the actual surfaces look rather complicated it's possible that the model is overfitting to noise here.

### Carbon-14 - Bomb-pulse corrected

Above we tried to compensate for the atmospheric Carbon-14 from the nuclear weapons testing by applying an AR(1) component. Another approach could be to modify the time series of measurements instead, detrending it. We load some atmospheric C-14 measurements that are in the same unit, and subtract a scaled version to attempt to make an approximately flat time series (save for any peaks due to e.g. marine discharges).

```{r}
data_C14 <- combine_data("C14", df_long_sardal, df_long_discharges, scale_data = FALSE)

bomb_pulse <- read_excel("~/RadiologyProject/sardal-radioactivity-analysis/S0033822221000953sup002.xls", col_names = FALSE, skip = 314)

bomb_pulse <- bomb_pulse %>%
  mutate(year = floor(.data[[names(bomb_pulse)[1]]])) %>%
  group_by(year) %>%
  summarise(F14C = mean(.data[[names(bomb_pulse)[4]]], na.rm = TRUE))

bomb_pulse <- bomb_pulse %>%
  bind_rows(
    bomb_pulse %>%
      slice_tail(n = 2) %>%
      mutate(slope = (F14C - lag(F14C)) / (year - lag(year))) %>%
      slice_tail(n = 1) %>%
      transmute(
        year = 2020,
        F14C = F14C + slope
      )
  )

bomb_pulse <- bomb_pulse %>%
  filter(year %in% data_C14$year)

data_C14$sardal <- data_C14$sardal - 0.68*bomb_pulse$F14C
plot(data_C14$sardal)

data_C14 <- data_C14 %>%
  add_lagged_matrices(n_max = MAX_LAG, n_min = MIN_LAG) %>%
  apply_weights()
```

We can now make a model again

```{r}
#| echo: true
#| output: false

if(FULL_RUN) {
  formula <- sardal ~
    te(sellafield, lag, k=4) +
    te(la_hague, lag, k=4)
  
  model_C14_Corrected_mvgam <- mvgam(formula = formula,
                           family = lognormal(),
                           data = data_C14,
                           trend_model = 'None',
                           burnin = 2000,
                           samples = 4000,
                           parallel = TRUE,
                           threads = 6,
                           noncentered = TRUE
                           )
  
  save(model_C14_Corrected_mvgam, file = "results/models/C14_Corrected.mvgam.Rdata")
}
```

```{r}
summary(model_C14_Corrected_mvgam)
plot_mvgam_smooth(model_C14_Corrected_mvgam, series = 1, smooth = 1)
plot_mvgam_smooth(model_C14_Corrected_mvgam, series = 1, smooth = 2)
plot(hindcast(model_C14_Corrected_mvgam))
plot_mvgam_resids(model_C14_Corrected_mvgam)
ppc(model_C14_Corrected_mvgam)
```

### Caesium-137

Now, caesium is a unique case, because the influence from the 1986 Chernobyl accident on Ceasium concentrations in the Nordics has been well studied. In particular, outflow from the Baltic sea is predicted to be the primary contributor of ceasium activity concentrations on the Swedish west coast. Thus, we must include it. We include here a Baltic sea caesium-137 simulation based on information from HELCOM (2009) *Radioactivity in the Baltic Sea, 1999-2006 HELCOM thematic assessment.* Balt. Sea Environ. Proc. No. 117, pages 21 and 22.

```{r}
simulate_baltic_cs137 <- function(data_Cs137) {
  
  # take a vector of years from the original dataset
  years <- sort(unique(data_Cs137$year))
  total_years <- length(years)
  # number of data points after the incident
  n_after <- length(years[years >= 1986])

  t <- seq_len(n_after) - 1

  # effective half lives used from HELCOM 2009 pp. 21-22
  k_1 <- log(2)/2.5 # 1986-1988
  k_2 <- log(2)/9 # 1990 - end
  y_0 <- 650 # approximate peak value Bq / m^-3
  
  # prefill empty array
  x <- array(5, total_years)
  
  # expoenential decay with piecewise half lives
  data <- y_0 * exp(- (k_1 * pmin(t, 3) + k_2 * pmax(0, t - 3)))
  
  # replace the tail (from 1986 to end) by sim data
  x[(total_years - n_after + 1):total_years] <- data
    
  # add noise
  set.seed(123)
  x <- x + rnorm(total_years, sd = 0.1 * mean(x))
  return(as.vector(x))
}
```

Adding baltic sea simulation to the data:

```{r}
data_Cs137 <- combine_data("Cs137", df_long_sardal, df_long_discharges, scale_data = FALSE)

baltic_Cs137 <- simulate_baltic_cs137(data_Cs137)

data_Cs137 <- data_Cs137 %>%
  group_by(series) %>%
  mutate(baltic = baltic_Cs137) %>%
  ungroup()

plot_isotope(data_Cs137)
```

Let's check correlation once more

```{r}
cor(diff(data_Cs137$sellafield), diff(data_Cs137$baltic))
```

We prepare the data as before

```{r}
data_Cs137 <- data_Cs137 %>%
  add_lagged_matrices(n_max = MAX_LAG, n_min = MIN_LAG) %>%
  apply_weights()
```

Note the extra terms below in the formula. It looks like we have a much bigger model, but this is the hierarchical model at play through applying the terms only when the weights matrix contains a 1. Anyways, let's fit

#### Model

```{r}
#| echo: true
#| output: false

if(FULL_RUN) {
  formula <- sardal ~
    te(sellafield, lag,
       k = 4) +
    te(baltic, lag,
       k = 4)
  
  model_Cs137_mvgam <- mvgam(formula = formula,
                           family = lognormal(),
                           data = data_Cs137,
                           burnin = 2000,
                           samples = 4000,
                           parallel = TRUE,
                           threads = 6,
                           noncentered = TRUE
                           )

  save(model_Cs137_mvgam, file = "results/models/Cs137.mvgam.Rdata")
}
```

We can make the same plots and summaries as before:

```{r}
summary(model_Cs137_mvgam)
mcmc_plot(model_Cs137_mvgam)
```

By p-values, it appears the model is finding predictive power for both Sellafield discharges and Baltic sea activity concentrations.

```{r}
plot_mvgam_smooth(model_Cs137_mvgam, smooth = 1)
plot_mvgam_smooth(model_Cs137_mvgam, smooth = 2)
```

The effect seems to be mainly around 1-2 years delay with a long tail of influence towards higher lags. The hypothesis that the majority of Caesium-137 after 1986 originates from Baltic sea outflow into the Kattegat has been described and supported in Mattsson et al. (2025) by effective half life calculations. Our simulated series of activity concentrations based on reported effective half lives seems to corroborate this hypothesis.

That Sellafield has predictive power in the model, could indicate that the peak in activity concentrations around the 1980s before the Chernobyl disaster is caused by the significant discharges from Sellafield in the late 1970s. The smooth shows a peak effect much later than initial arrival time, closer to 6-7 years.

```{r}
plot(hindcast(model_Cs137_mvgam))
plot(hindcast(model_Cs137_mvgam), series = 2)
```

The trends look slightly different. We can see what the difference is:

```{r}
plot(model_Cs137_mvgam, type = "re")
```

It seems like the uncertainties are rather large and we cannot say anything conclusive from this model whether F. Serratus and F. Vesiculosus have different uptake of Caesium-137 in the Swedish west coast marine environment.

And the diagnostics:

```{r}
plot_mvgam_resids(model_Cs137_mvgam)
ppc(model_Cs137_mvgam)
```

### Cobalt-60

```{r}
data_Co60 <- combine_data("Co60", df_long_sardal, df_long_discharges, scale_data = FALSE)
plot_isotope(data_Co60)
```

```{r}
cor(diff(data_Co60$barseback), diff(data_Co60$la_hague))
cor(diff(data_Co60$barseback), diff(data_Co60$winfrith))
cor(diff(data_Co60$winfrith), diff(data_Co60$la_hague))
```

Now, for Cobalt-60 we have rather limited data available, and in order to get a useful model, one cannot have too many parameters compared to number of datapoints. Thus, since the signal for Sellafield is rather weak compared to the rest we ignore it. We prepare the data as before

```{r}
data_Co60 <- data_Co60 %>%
  add_lagged_matrices(n_max = MAX_LAG, n_min = MIN_LAG) %>%
  apply_weights()
```

Let's fit as before

#### Model

```{r}
#| echo: true
#| output: false

if(FULL_RUN) {
  formula <- sardal ~
    te(la_hague, lag,
       k = 4) +
    te(barseback, lag,
       k = 4)

  model_Co60_mvgam <- mvgam(formula = formula,
                           family = lognormal(),
                           trend_model = 'None',
                           data = data_Co60,
                           burnin = 2000,
                           samples = 4000,
                           parallel = TRUE,
                           threads = 6,
                           noncentered = TRUE
                           )

  save(model_Co60_mvgam, file = "results/models/Co60.mvgam.Rdata")
}
```

We can make the same plots and summaries as before:

```{r}
summary(model_Co60_mvgam)
mcmc_plot(model_Co60_mvgam)
```

```{r}
plot_mvgam_smooth(model_Co60_mvgam, smooth = 1)
plot_mvgam_smooth(model_Co60_mvgam, smooth = 2)
```

```{r}
plot(hindcast(model_Co60_mvgam), series = 1)
plot(hindcast(model_Co60_mvgam), series = 2)
```

```{r}
plot(model_Co60_mvgam, type = "re")
```

And the diagnostics:

```{r}
plot_mvgam_resids(model_Co60_mvgam)
ppc(model_Co60_mvgam)
```

### Iodine-129

For Iodine-129, Mattsson et al. (2025) report significant seasonal differences between summer and winter data. Because this is a significant effect, and we have the collection season available as data, we chose here to treat F. Serratus in summer and winter as two separate 'species'. We drop the mean measurement which is denoted as just F. Serratus:

```{r}
data_I129 <- combine_data("I129", df_long_sardal, df_long_discharges, scale_data = FALSE)

plot_isotope(data_I129)

cor(diff(data_I129$sellafield), diff(data_I129$la_hague))

data_I129 <- droplevels(data_I129[!data_I129$series == 'I129_F_Serratus',])

data_I129 <- data_I129 %>%
  add_lagged_matrices(n_max = MAX_LAG, n_min = MIN_LAG) %>%
  apply_weights()
```

#### Model

Let's fit as before

```{r}
#| echo: true
#| output: false

if(FULL_RUN) {
  formula <- sardal ~
    s(series, bs = "re") +
    te(sellafield, lag, k = 4, by=weights_ss) +
    te(sellafield, lag, k = 4, by=weights_sw) +
    te(la_hague, lag, k = 4, by=weights_ss) +
    te(la_hague, lag, k = 4, by=weights_sw)
  
  model_I129_mvgam <- mvgam(formula = formula,
                            family = lognormal(),
                            data = data_I129,
                            trend_model = 'None',
                            burnin = 2000,
                            samples = 4000,
                            parallel = TRUE,
                            threads = 6,
                            noncentered = TRUE,
                            priors = priors
                             )

  save(model_I129_mvgam, file = "results/models/I129.mvgam.Rdata")
}
```

We can make the same plots and summaries as before:

```{r}
summary(model_I129_mvgam)
mcmc_plot(model_I129_mvgam)
```

```{r}
plot_mvgam_smooth(model_I129_mvgam, smooth = 1)
plot_mvgam_smooth(model_I129_mvgam, smooth = 2)
plot_mvgam_smooth(model_I129_mvgam, smooth = 3)
plot_mvgam_smooth(model_I129_mvgam, smooth = 4)
```

```{r}
plot(hindcast(model_I129_mvgam), series = 1)
plot(hindcast(model_I129_mvgam), series = 2)
```

```{r}
plot_mvgam_randomeffects(model_I129_mvgam)
```

And the diagnostics:

```{r}
plot_mvgam_resids(model_I129_mvgam)
ppc(model_I129_mvgam)
```

## Generative AI statement

Generative AI (GAI) was **not** **used** for text generation anywhere in this document or the final report.

GAI (Mistral 3, GPT-4, GPT-5) **has been used** for debugging and for creating and modifying visualization code (e.g. code for plot titles, labels, fonts, colour schemes).

GAI was **not** **used** for "mission-critical" code, such as the loading/wrangling of data and the specification of the model.

GAI was **not** **used** for generation or interpretation of the results.

All output from GAI that has been used was reviewed and edited to ensure that any output used in the final document was correct and achieved its intended purpose. The author takes full responsibility for the contents of this document.
